---
title: "Data Wrangling Exercise 2"
author: "Carlee Price"
date: "May 28, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Wrangling Exercise 2

0: Load the data into RStudio

```{r}
titanic = read.csv("titanic_original.csv")
```

1: Port of Embarkation

```{r}
levels(titanic$embarked)[1] <- "S"
```

2: Age

```{r}
#calculate the mean age
x <- mean(titanic$age, na.rm = TRUE)
#& use it to populate missing values
titanic$age <- replace(titanic$age, is.na(titanic$age), x)
```

Other strategies for populating missing values: other aggregate statistics for the group, such as median.  Or one could use a summary statistic for a narrower group within the dataframe.  Average age for women, by example, may be different than for men.  Or passengers who were assigned a cabin versus those who were not.

3: Lifeboat

```{r}
levels(titanic$boat)[1] <- "None"
```

4: Cabin

It does *not* make sense here to populate the empty fields with a value.  Missing value means the passenger was cabinless, or travelling in steerage.  

```{r}
patt <- grepl("[A-Z]", titanic$cabin)
titanic$has_cabin_number <- as.numeric(patt)
```

6. Prepare for submission to github

```{r}
write.csv(titanic, file = "titanic_clean.csv")
```
